\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-04 chapters/bci.tex
Rewrite sota

\todo{comments liesel paradigms}
\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}
\todo{use terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todofc{chapters should look more like full papers}
\todo{uniform vsa condition vsa setting}
\todo{acronyms written in full every new chapter and in captions}
\the\textwidth

People with severe physical impairment due to
neurodegenenerative disease, stroke, traumatic brain or spine injury
become entirely reliant on caregivers or family members in their day to day
routine.
Wouldn't it be great if we could help them regain some quality of life by
making them more independent?

@@

Because of the desirable high temporal resolution~\cite{Easttom2021} and
practicality of electrophysiology, we will focus only on methods to record
electromagnetical activity originating from neuronal action potentials and
postsynaptic potentials.
The magnetical field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, the necessary
equipment is rather expensive and impractical.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
lab conditions.

As a consequence, the \ac{bci} field relies heavily on the
electrical activity of the brain.
The \emph{invasiveness} of the technology forms an
important concept in determining its suitability for a given application and
patient.
As illustrated in Figure~\ref{fig:bci/recording}, invasive electrodes can often
measure a very specific brain region and can result in better signal quality, at the cost of the risks
involved with brain surgery.

@@

@@

The major drawback of \ac{eeg} is its poor spatial resolution.
The electrodes are large and sit on the outside of the scalp, relatively far
away from the cortex.
Furthermore, the layers of skull and cerebrospinal fluid in between can
attenuate the signal and cause volume conduction, which results in electrodes
measuring a mixture of activity from sources elsewhere in the brain.
This also has a negative effect on spectral resolution.
The noise problem mentioned earlier is also prominently present in \ac{eeg}
recordings.
In addition to picking up brain activity other than that of the region of
interest, \ac{eeg} also records other artifacts, from the environment, the power
supply, nearby electrical equipment, or the user's muscle movements.
Combined, this means \ac{eeg} is inherently noisy.
\todo{citations from 'Decoding covert speech from eeg-a cromprehensive review
EEG has lower signal-to-noise ratio (SNR) than the other
modalities. It is almost always corrupted by artifacts such as
muscular artifacts (Eberle et al., 2012; Liu, 2019).
• EEG has limited spectral and spatial resolution (Peled et al.,
2001; Lakshmi et al., 2014).
• Recording EEG for longer duration is challenging since the
conductive gel or the saline solution applied for reducing the
electrode impedance dries up over time, thus increasing the
electrode impedance (Guger et al., 2012; Xu et al., 2018a).
• A trained personnel is required for placing the EEG
electrode cap}

@@

\section{Paradigms}
\label{sec:bci/paradigms}
\todohsm{bring more structure, more subtitles}

@@

\fullpagefig{%
  \input{figures/bci/paradigms.tikz.tex}
}{%
  Overview of \ac{bci} paradigms.
}{%
  A \ac{bci} paradigm defines which brain activity will be used for control.
  They can be categorized in active, reactive and passive
  paradigms, depending on their reliance on external stimulation and active
  participation required from the user.
  Adapted from~\textcite{Muehl2014}.
}{%
  fig:bci/paradigms
}
\todo{check permissions}
\todo{above Zander and Kothe was mentioned.}

Paradigms that require little active participation of the user like
redirecting attention or initiating imagined actions are classified as \emph{passive
\ac{bci}}.
From the user's point of view, this concept is probably the most attractive,
since their intentions or cognitive state state would be inferred from just
their brain activity.\todohsm{in "passive bci", the user is not supposed to express "intentions". ragher start with active "}
However, without active user participation, it can be hard to establish reliable
communication or control.
Therefore, passive \ac{bci} are currently more suited to tasks like
emotion or affect monitoring~\cite{Torres2020,Libert2019}, workload
monitoring~\cite{Zanetti2021} or, more dependent on stimulation, error
detection~\cite{SiMohammed2020}.\todo{mention neurofeedback}
Furthermore, active paradigms\todo{"active" is used in a fuzzy way: active BCI is self-paced, reactive is in response to a cued event. Further down that becomes more clear.
The "active participation": is needed in both cases as shown in Fig. 1.3.
I suggest to say active participation} make it easier to quickly collect training data for
supervized machine learning since the conditions can be more easily controlled
by the user or the stimulation.

Paradigms with high active participation can be split up in \emph{active
\ac{bci}} and \emph{reactive \ac{bci}}.
Active \ac{bci} paradigms encode endogenous activity initiated by the user,
such as imagined movement or imagined speech.
These tasks can encode complex information, but current non-invasive
communication methods often limit the considered domain to a few movement
directions to control a cursor, or a few words~\cite{Panachakel2021}.
Invasive recordings are necessary for decoding more complex encoded
like natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
symbols~\cite{Willett2021} or sign language~\cite{Branco2017}.
Active paradigms are also subject to large inter-subject variability due to the
complexity of the performed tasks.
Imagined speech or movements can be performed in a multitude of ways, which
can each have different neural representations for separate individuals.
This can make it hard to adapt the \ac{bci} to specific individuals, causing
poor performance and giving rise to the concept of \ac{bci}
illiteracy~\cite{Allison2010}\footnote{%
Recently, the concept of BCI illiteracy is increasingly being seen as outdated.
Instead of attributing poor performance to the user's inherent limitations,
critiques point out that the issue lies more with the design of the
systems themselves. BCI systems often fail to account for individual
variability in brain anatomy and cognitive strategies. As a result, what has
been labeled as "illiteracy" may, in fact, be a reflection of systems that are
not flexible enough to adapt to different users. This perspective
shifts the responsibility to the individual, rather than recognizing that
better, more adaptive systems could overcome many of these performance
challenges, thus standing in the way of progress.
Moreover, the term ``illiteracy'' itself carries negative
connotations, implying a user deficiency that reinforces normative assumptions,
when in reality, a more inclusive approach to design could enhance BCI
accessibility for a wider range of individuals. This evolving view calls for a
reframing of the problem, focusing on improving the adaptability of BCIs rather
than labeling users as illiterate~\cite{Becker2022,Thompson2019}.
}.
\todo{user training more than with reactive, task instruction}

Reactive \ac{bci} take another approach, by providing a discrete set of
sensory stimulations towards which the user's attention can directed.
Compared to active \ac{bci}, reactive paradigms usually require more effort\todohsm{is this the case?}
and can induce fatigue due to the constant stimulation.
These paradigms are also somewhat less intuitive compared to e.g.\ speech, and
their expressive power is limited by how many different stimuli can effectively
presented and attended within a reasonable timeframe.
Nevertheless, reactive \ac{bci} work well with \ac{eeg} recordings, and, most
importantly, they work for most people~\cite{Allison2010a,Edlinger2014}.

@@

The major visual paradigms rely on
\ac{ssvep}, \ac{cvep}, oddball and \ac{mvep} brain responses.
In \ac{ssvep}, information is modulated by the frequency of a stimulus attended
among many that each flicker continuously at different
frequencies~\cite{Chen2021}.
In \ac{cvep}, the stimuli flicker instead with distinct on-of patterns, which can
be correlated with the brain activity to retrieve the attended stimulus.
Instead of stimulating all possible selections at once, the oddball paradigm
stimulates them one by one with single flashing.\todo{reference}
Since the times of stimulation of each target is known, and time-points at
which an attention signature is detected to selected targets.
\Ac{mvep} is usually similarly time modulated, but stimuli make sudden
movements in specific directions instead of flashing.
Information is then also carried by movement
direction~\cite{Libert2021a,Libert2022}.

@@

\section{Preprocessing}%
\label{sec:bci/preprocessing}
\todo{preprocessing}
EEG preprocessing is a critical step in brain-computer interface (BCI) systems, as it
significantly improves the quality of the recorded data for subsequent analysis and
classification. Common preprocessing techniques in BCI systems include
rereferencing, band-pass and notch filtering, and \ac{ica}.

@@

Band-pass filtering is then usually applied to isolate the frequency ranges of interest by
removing frequencies that are too high or too low to contain relevant neural signals.
Band-pass filters eliminate slow drifts and high-frequency noise, such as
muscle artifacts or environmental electrical interference, improving the
\ac{snr}.
Additionally, notch filtering can be  used to remove strong power line
interference\footnote{Typically at 50Hz in Europe}.
This interference can heavily contaminate EEG recordings and even leak through
when outside the passband of a previous filter.
Notch filtering efficiently attenuates this narrowband noise without affecting
the surrounding frequencies.

@@

@@

@@

\todo{four pane figure:
  photo of someone using a P300 bci
  P300 matrix interface
  simplified figure of ERP components
  full spatiotemporal ERP response
}

\subsection{The \acf{erp}}
\label{sec:bci/oddball/erp}

The time-domain response elicited immediately after the intensification of a
stimulus is known as the visual \ac{erp}~\cite{Luck2014}.
An \ac{erp} is a waveform consisting of multiple components.
Some of these components are modulated by whether the stimulus was an
oddball or not.
These components appear as positive peaks and negative troughs in the \ac{erp}
waveform.
They follow a naming scheme based on polarity (Negative or Positive) and
latency (e.g., N2 or N200 for a negative component after $\pm200$ ms).
The component most prominently modulated by attention is called the P3
(positive deflection after $\pm300$ ms).\todo{add footnote about how ms naming
is maybe not suited, especially since this work focuses on the fact that there
is a lot of variability in the timings for visual potentials. the short naming
scheme builds on 'the first negative component', 'the first positive
component', and for the visual oddball erp this ordening more or less coincides
with the timings.}
Therefore, oddball paradigms are frequently referred to as P300 paradigms.
However, due to the equally important contributions of other \ac{erp}
components in decoding, as we will show later, this could be considered a
misnomer.
Thus, we will adhere to the oddball paradigm naming.

@@

\subsection{Preprocessing \& feature extraction}
In addition to (ref) epoching
baseline correction, but not in decoding (cite)

\subsection{Decoding ERPs}

Usually, snr low and multiple trials necessary

tLDA, beamforming
Linear decoding techniques

@@

\section{The gaze-dependence problem}

>>>>

\todo{comments liesel paradigms}
\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}
\todo{use terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todofc{chapters should look more like full papers}
\todo{uniform vsa condition vsa setting}
\todo{acronyms written in full every new chapter and in captions}
\todo{search for ??}

People with \ac{sspi} due to
neurodegenenerative disease, stroke, traumatic brain or spine injury
become entirely reliant on caregivers or family members in their day to day
routine.
Wouldn't it be great if we could help them regain some quality of life by
making them more independent?

@@

Because of the desirable high temporal resolution~\cite{Easttom2021} and
practicality of electrophysiology, we will focus only on methods to record
electromagnetical activity originating from neuronal action potentials and
postsynaptic potentials.
The magnetical field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, the necessary
equipment is rather expensive and impractical.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
lab conditions.
As a consequence, the \ac{bci} field relies heavily on the
electrical activity of the brain.

\subsection{Invasive electrophysiology}
The \emph{invasiveness} of the technology forms an
important concept in determining its suitability for a given application and
patient.
As illustrated in Figure~\ref{fig:bci/recording}, invasive electrodes can often
measure a very specific brain region and can result in better signal quality, at the cost of the risks
involved with brain surgery.

@@

\subsection{\Acl{eeg}}

@@

The major drawback of \ac{eeg} is its poor spatial resolution~\cite{Ferree2001}.
Since they sit on the outside of the scalp electrodes are large and distance
relative to the neurons in the cortex they ought to measure.
Furthermore, the layers of skull and cerebrospinal fluid in between can
attenuate the signal and cause volume conduction~\cite{Broek1998}, which results in electrodes
measuring a mixture of activity from sources elsewhere in the brain.
This also has a negative effect on spectral resolution.
The noise problem mentioned earlier is also prominently present in \ac{eeg}
recordings.
In addition to picking up brain activity other than that of the region of
interest, \ac{eeg} also records other artifacts, from the environment, the power
supply, nearby electrical equipment, or the user's muscle
movements~\cite{Urigueen2015}.
Combined, this means \ac{eeg} is inherently noisy.

@@

\section{Paradigms}
\label{sec:bci/paradigms}

\subsection{Active \& passive \ac{bci}}

@@

\fullpagefig{%
  \input{figures/bci/paradigms.tikz.tex}
}{%
  Overview of \ac{bci} paradigms.
}{%
  A \ac{bci} paradigm defines which brain activity will be used for control.
  They can be categorized in active, reactive and passive
  paradigms, depending on their reliance on external stimulation and active
  participation required from the user.
  Adapted from~\textcite{Muehl2014}.
}{%
  fig:bci/paradigms
}

Paradigms that require little active participation of the user like
redirecting attention or initiating imagined actions are classified as \emph{passive
\ac{bci}}.
From the user's point of view, this concept is probably the most attractive,
since their cognitive state state would be inferred from just
their brain activity.
However, without active user participation, it can be hard to establish reliable
communication or control.
Therefore, passive \ac{bci} are currently more suited to tasks like
emotion or affect monitoring~\cite{Torres2020,Libert2019, Muehl2014}, workload
monitoring~\cite{Zanetti2021} or, more dependent on stimulation, error
detection~\cite{SiMohammed2020}.
It also has clinical applications in the form of neurofeedback~\cite{Hammond2011}.
Nevertheless, active participation facilitates training data collection for
supervised machine learning since the conditions can be more easily controlled
by the user or the stimulation.

Paradigms requiring high active participation can be split up in \emph{active
\ac{bci}} and \emph{reactive \ac{bci}}.
Active \ac{bci} paradigms encode endogenous activity initiated by the user,
such as imagined movement or imagined speech.
These tasks can encode complex information, but current non-invasive
communication methods often limit the considered domain to a few movement
directions to control a cursor, or a few words~\cite{Panachakel2021}.
Invasive recordings are necessary for decoding more complex encoded
like natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
symbols~\cite{Willett2021} or sign language~\cite{Branco2017}.
Active paradigms are also subject to large inter-subject variability due to the
complexity of the performed tasks.
They might require extensive user training and are sensitive to
correct task instruction.
Imagined speech or movements can be performed in a multitude of ways, which
can each have different neural representations for separate individuals.
This can make it hard to adapt the \ac{bci} to specific individuals, causing
poor performance and giving rise to the concept of \ac{bci}
illiteracy~\cite{Allison2010}\footnote{%
Recently, the concept of BCI illiteracy is increasingly being seen as outdated.
Instead of attributing poor performance to the user's inherent limitations,
critiques point out that the issue lies more with the design of the
systems themselves. BCI systems often fail to account for individual
variability in brain anatomy and cognitive strategies. As a result, what has
been labeled as `illiteracy' may, in fact, be a reflection of systems that are
not flexible enough to adapt to different users. This perspective
shifts the responsibility to the individual, rather than recognizing that
better, more adaptive systems could overcome many of these performance
challenges, thus standing in the way of progress.
Moreover, the term `illiteracy' itself carries negative
connotations, implying a user deficiency that reinforces normative assumptions,
when in reality, a more inclusive approach to design could enhance BCI
accessibility for a wider range of individuals. This evolving view calls for a
reframing of the problem, focusing on improving the adaptability of BCIs rather
than labeling users as illiterate~\cite{Becker2022,Thompson2019}.
}.

\subsection{Reactive \ac{bci}}
Reactive \ac{bci} take another approach, by providing a discrete set of
sensory stimulations towards which the user's attention can directed.
Compared to active \ac{bci}, reactive paradigms can more easily induce fatigue
due to the constant stimulation.
These paradigms are also somewhat less intuitive compared to e.g.\ speech, and
their expressive power is limited by how many different stimuli can effectively
presented and attended within a reasonable timeframe.
Nevertheless, reactive \ac{bci} work well with \ac{eeg} recordings, and, most
importantly, they work for most people~\cite{Allison2010a,Edlinger2014}.

@@

The major visual paradigms rely on
\ac{ssvep}, \ac{cvep}, oddball and \ac{mvep} brain responses.
In \ac{ssvep}, information is modulated by the frequency of a stimulus attended
among many that each flicker continuously at different
frequencies~\cite{Chen2021}.
In \ac{cvep}, the stimuli flicker instead with distinct on-of patterns, which can
be correlated with the brain activity to retrieve the attended
stimulus~\cite{Sun2022}.
Instead of stimulating all possible selections at once, the oddball paradigm
stimulates them one by one with single flashes~\cite{Pan2022}.
Since the times of stimulation of each target is known, and time-points at
which an attention signature is detected to selected targets.
\Ac{mvep} is usually similarly time modulated, but stimuli make sudden
movements in specific directions instead of flashing.
Information is then also carried by movement
direction~\cite{Libert2021a,Libert2022}.

@@

\section{Preprocessing}%
\label{sec:bci/preprocessing}
EEG preprocessing is a critical step in brain-computer interface (BCI) systems, as it
significantly improves the quality of the recorded data for subsequent analysis and
classification. Common preprocessing techniques in BCI systems include
rereferencing, band-pass and notch filtering, and \ac{ica}.

@@

Band-pass filtering is then usually applied to isolate the frequency ranges of interest by
removing frequencies that are too high or too low to contain relevant neural signals.
Band-pass filters eliminate slow drifts and high-frequency noise, such as
muscle artifacts or environmental electrical interference, improving the
\ac{snr}.
Additionally, notch filtering can be  used to remove strong power line
interference\footnote{Typically at 50Hz in Europe.}.
This interference can heavily contaminate EEG recordings and even leak through
when outside the passband of a previous filter.
Notch filtering efficiently attenuates this narrowband noise without affecting
the surrounding frequencies.

@@

\subsection{Design principles}

@@

\subsection{Implemetation \& evaluation}

@@

\begin{figure}[t]
  \begin{subfigure}[t]{.45\textwidth}
  \includegraphics[width=\textwidth, height=\textwidth]{example-image-a}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/bci/p300.tikz.tex}
    \caption{The P300 matrix speller interface. Entire rows or columns flash at once.}
  \end{subfigure}

  \begin{subfigure}[t]{.45\textwidth}
    \input{figures/bci/erp.tikz.tex}
    \caption{The idealized \ac{erp} evoked by an attended stimulation.}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth, height=0.6180\textwidth]{example-image-b}
  \end{subfigure}
\end{figure}

\todo{four pane figure:
  photo of someone using a P300 bci
  full spatiotemporal ERP response
}

\subsection{The \acf{erp}}%
\label{sec:bci/oddball/erp}

The time-domain response elicited immediately after the intensification of a
stimulus is known as the visual \ac{erp}~\cite{Luck2014}.
An \ac{erp} is a waveform consisting of multiple components.
Some of these components are modulated by whether the stimulus was an
oddball or not.
These components appear as positive peaks and negative troughs in the \ac{erp}
waveform.
They follow a naming scheme based on polarity (Negative or Positive) and
latency (e.g., N2 or N200 for a negative component after $\pm200$ ms).
The component most prominently modulated by attention is called the P3
(positive deflection after $\pm300$ ms)
some \ac{bci}\footnote{In \ac{erp} analysis,
components are sometimes referred to with the timing around which they occur
(e.g.\ N170, negative after 170ms), or by their rank in order of occurrence
(N1, the first negative component).
The timing nomenclature is based on average latencies for neurotypical
individuals, but is seldom correct for specific \ac{bci} user due to the large
variability across \ac{erp} stimulation paradigms and subjects and can also
depend on the user's pathology.
Furthermore, as we will see later, this work focuses specifically on the
intra-session variability in \ac{erp} latency, explicitly assuming the latencies are not
fixed.
Therefore, we adhere to the ranking nomenclature.
Fortunately, for the visual oddball paradigm, the rankings do roughly
correspond to their expected timings, i.e.\ P1=P100, N1=N100, etc.}.
Therefore, oddball paradigms are frequently referred to as P3(00) paradigms.
However, due to the equally important contributions of other \ac{erp}
components in decoding, as we will show later, this could be considered a
misnomer.
Thus, we will adhere to the oddball paradigm naming.

@@

\subsection{Preprocessing \& feature extraction}
For \ac{erp} analysis, the \ac{eeg} is usually band-pass filtered between 0.5Hz
and 16Hz.
Re-referencing can be done to average mastoids to highlight the contribution of
the P3 component.
In addition to the steps outlined in section~\ref{sec:bci/preprocessing},
the signal is usually cut into \emph{epochs} for \ac{erp} analysis.
These are time windows surrounding the stimulus event.
They usually include some baseline interval before the stimulus, which can be
used to normalize the response.

The epochs contain values representing the voltage of a specific channel relative to the
reference at a specific time relative to the stimulus.
In \ac{erp} classification, they are usually directly used as features, by
concatenating all channels in the matrix format epoch $X$ into one feature
vector s.t.\ $\mat{x} = \mat\vec(X)$.

\subsection{Decoders}
Several techniques are suited to decode~\ac{erps}.
Below are some methods that can achieve state-of-the art performance and are
relevant to this work.

\subsubsection{\Acs{lcmv} beamforming}
Linear decoding techniques such as the \ac{lcmv}
beamformer~\cite{Wittevrongel2016} generally work well for
classifying \ac{erp} epochs.
\Ac{lcmv} beamforming learns a weight vector $\mat{w}$ from a set of epochs that can
be multiplied with the feature vector of a new epoch to obtain a prediction
score.
The weights $\mat{w}$ are obtained by minimizing the variance of the output
\begin{equation}
  \argmin_\mat{w}\mat{w}^\intercal\mat{C}
	\mat{w}^\intercal
\end{equation}
under the linear constraint
\begin{equation}
	\mat{a}^\intercal\mat{w} = 1
\end{equation}
with activation pattern $a$ the difference between the average of the attended epochs and the
average of the non-attended epochs in the training data.
This is then solved by  method of Lagrange multipliers as
\begin{equation}
	\mat{w} =
  \frac{\mat{C}^{-1}\mat{a}^\intercal}
  {\mat{a}\mat{C}^{-1}\mat{a}^\intercal}
\end{equation}
with $\mat{C}$ the background covariance.

\subsubsection{\Ac{lda}}
\Ac{lda} is a more well-known method that is, int fact, nearly equivalent to
\ac{lcmv} beamforming~\cite{Treder2016} under certain conditions.
It assumes normally distributed data with a different mean per class but a
common covariance.
Similar to \ac{lcmv}-beamforming, \ac{lda} learns a weight vector $\mat{w}$
that satisfies
\begin{equation}
  \argmax_\mat{w}\frac{\mat{w}^\intercal\mat{S}_\text{b}\mat{w}}{\mat{w}^\intercal\mat{S}_\text{w}\mat{w}}
\end{equation}
The between-class scatter matrix $\mat{S}_\text{b}$ models the variability
between the expected responses, the within-class scatter matrix
$\mat{S}_\text{w}$ the expected noise.
The optimization criterion corresponds to finding a projection that maximizes
the spread between classes (signal) while minimizing the spread within classes
(noise).
$mat{w}$ can be obtained as eigenvector of
$\mat{S}_\text{w}^{-1}\mat{S}_\text{b}$ corresponding to the largest eigenvalue.

There is one one caveat with these linear methods, however:
the amount of values ($=\text{nb. of channels}\times\text{nb. of sampled time points}$)
is usually relatively high compared to the training dataset, since calibration
time should be kept minimal.
Therefore, strong regularization should be applied, such as covariance/scatter
shrinkage.
Regularization can also be performed by retaining structure in the data, such
as is the case for the multilinear methods presented by~\textcite{Lotte2018}.

\subsubsection{\Ac{tlda}}

Another example of this is \ac{tlda}~\cite{Sosulski2022}, which assumes temporal
stationarity on the background noise to regularize the between-class covariance
matrix of \ac{lda}
\begin{equation}
  \mat{S}_\text{w} = \begin{bmatrix}
s_0 & s_1 & s_2 & \cdots & s_{n-1} \\
s_1 & s_0 & s_1 & \cdots & s_{n-2} \\
s_2 & s_1 & s_0 & \cdots & s_{n-3} \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
s_{n-1} & s_{n-2} & s_{n-3} & \cdots & s_0
\end{bmatrix}
\end{equation}
This method is simple yet effective reaching state-of-the-art decoding
performance.

\subsubsection{Riemannian geometry}

Alternative methods that currently work well include
non-linear methods include Riemannian geometry~\cite{Barachant2014}.
Methods leveraging Riemannian geometry use a different feature extraction
method.
They do not use the raw epoch values, but rather operate on the spatial
covariance of an epoch.
These covariance matrices $C_i$ are symmetric and should have only positive
eigenvalues, hence they lie on the Riemannian manifold of symmetric positive
definite matrices, which defined distance metric
\begin{equation}
  \delta_\text{R}\left(\mat{C}_1,\mat{C}_2\right)
  =\left\lVert\log\left(\mat{C}_1^{-\frac{1}{2}} \mat{C}_2 \mat{C}_1^{-\frac{1}{2}}\right)\right\rVert
\end{equation}
This geodesic distance metric is naturally suited for this type of data and has
some desirable properties, like invariance to linear transformations, which
result in a robustness particularly useful for \ac{bci}
applications~\cite{Barachant2011}.

With this distance defined, the epochs can than be either directly be
classified on the Riemannian manifold using a minimum distance to mean
classifier, or they can be projected to the tangent space.
This is a euclidean space that locally approximates the Riemannian distance,
which allows us to apply regular machine learning methods like an SVM or
\ac{lda}.
The tangent space is obtained
To enhance classifications of \ac{erps} which rely on time domain information
which would otherwise be lost in the spatial representation of the extracted
feature, the class averages can be concatenated as additional channels to the
epochs, embedding the temporal structure in the feature~\cite{Barachant2014}.
To improve performance even further, the \acp{erp} can be decomposed by a method
like XDAWN before calculating the extended covariances~\cite{Li2020}.

These decoders define the last building blocks of our visual oddball \ac{bci}.
Yet, the overview until now has been rather theoretical.
Interesting and complex challenges arise when applying these concepts outside
of the lab in augmented and assistive communication technologies for patients
with \acp{sspi}.

@@

### 2024-10-04 chapters/conclusion.tex
Rewrite sota

\section{Recommendations}
\subsection{Working with patients}
gaze independent visual, can be good current existing solution
take into accounrecommendations of \textcite{FriedOken2020}

>>>>

\section{Recommendations}

this work focused on classification methods, and solving problem through
decoding

classification performance is not everything

ITR, comfort and user satisfaction in on-line operation is the ultimate goal

also do advance calculations of what the information transfer rate would be
given reasonable accuracy, so that you can go in with a reasonable interface
that has state-of-the-art ITR

classification performance is convenient since you can do a single data
collection and iterate offline.
this constricts you to decoder development
it is a better approach to only go there after optimizing the interface
experimentally and with patients with an existing
decoder, and only moving there once the options for progress are depleted

performance and abilities are probably so user dependent that it is not said that
findings from a population study with healthy controls will hold for a given
user with \ac{sspi}.
Incremental gains in performance in decoding for a specific paradigm will be
outweighed by how well that paradigm is adapted to the user, factoring in,
skills and abilities.

best approach should be everything working together, such as is advocated
by~\textcite{Pan2022} and \textcite{Fouad2020}

instead of decoder > data > patients > online
go other way
online > patients > data > decoder
this is unfortunately also the least practical way to do it for a 4-year
research project.

It is therefore important that research labs keep a working, long-term collaboration with
patient centers such that new projects can immediately be tested by the target
population, and that patients with bci experience can compare multiple systems
and guide development
The fact that a user has previous bci experience is often seen as a confounding
factor, while it is a strength that can help you gain insight into what works
and doesn't work for them, and where research efforts should be focused.

also maintain a working, on-line system that can be iterated upon

in a phase that is heavily dependent on user experimentation

not best approach for all research projects, but
Visual oddball bci field is advanced enough to do this

\subsection{Working with patients}
gaze independent visual, can be good current existing solution
take into accounrecommendations of \textcite{FriedOken2020}
